Modelling the emergence of whisker barrels

### 2020-07-14 paper/briefpaper/barrels_proof.tex
Adds the all-black final version

% Some document-defined commands
\newcommand*\dif{\mathop{}\!\mathrm{d}}

\input{sebcolour.tex}

% Miscellaneous changes
\newcommand{\cmnt}[1]{\textcolor{colcmnt}{#1}}
% Justify gamma/interaction parameter choice. major point 1.
\newcommand{\MPone}[1]{\textcolor{colmpone}{#1}}
% Blue changes are major point 2 (competition mechanism) and other, general
% uncategorised changes.
\newcommand{\MPtwo}[1]{\textcolor{colmptwo}{#1}}
% Parameter search. Major point 3
\newcommand{\MPthreePar}[1]{\textcolor{colmpthreepar}{#1}}
% Sensitivity analysis. Major point 3
\newcommand{\MPthreeSens}[1]{\textcolor{colmpthreesens}{#1}}
% More predictions. Major point 3
\newcommand{\MPthreePred}[1]{\textcolor{colmpthreepred}{#1}}
% Justify competition:
\newcommand{\mpfive}[1]{\textcolor{colmpfive}{#1}}
% Minor point 4:
\newcommand{\mpfour}[1]{\textcolor{colmpfour}{#1}}
% Minor point 6:
\newcommand{\mpsix}[1]{\textcolor{colmpsix}{#1}}
% Describe the metrics
\newcommand{\metrics}[1]{\textcolor{colmetrics}{#1}}

\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}
% Nice norms:
\usepackage{mathtools}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}
% Degrees symbol:
\usepackage{textcomp}

@@

Barrel formation requires afferent input from whisker stimulation and thalamic
calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex
network of axon guidance molecules such as ephrin-A5 and A7 and adhesion
molecules such as cadherin-6 and 8
\citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}.  This network
is orchestrated by interactions between morphogens Fgf8 and Fgf17 and
transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients \mpfour{spanning the cortical sheet} that mark orthogonal axes
and can be manipulated to stretch,
shrink, shift, and even duplicate barrels
\citep{assimacopoulos_fibroblast_2012}.

The barrel boundaries form a Voronoi tessellation \citep{senft_mouse_1991}
(Fig.\,\ref{fig:main}A), suggesting that barreloid topology is preserved in
the projection of thalamocortical axons into the cortex, and that a barrel
forms by lateral axon branching from an initial center-point that ceases upon
contact with axons branching from adjacent centers.  However, the assumption
of pre-arranged center-points is difficult to resolve with the observation
that axons arrive in the cortical plate as an undifferentiated bundle,
\emph{prior} to barreloid formation \citep{agmon_organized_1993}.  \mpsix{In
  mice, axons from the trigeminal ganglion arrive in the principal division of
  the trigeminal nucleus (PrV) at E12, then axons from the PrV arrive
  in the ventroposteromedial nucleus of the thalamus (VPM) at E17, then axons
  from the VPM arrive in the cortical plate at E18/P0. Distinct
  whisker-related clusters then become apparent in the PrV at P0-P1, in the
  VPM at P2-P3, and in the cortex at P3-P5}
\citep{erzurumlu_development_2012,sehara_neuronal_2011}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the \cmnt{fraction
  of occupied synapses} $c_i(x,t)$ and \cmnt{the density of} axon branches
$a_i(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for
$N$ thalamocortical projections indexed by $i$. The model was derived from the
assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally
coupled. Extending the original 1D model to simulate arealisation on a 2D
cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total \cmnt{fraction} of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise, \cmnt{$c_i(\mb{x},t)$} increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \MPtwo{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). Here $\chi_i=0$ is a
placeholder. The second term on the right represents the coupling between axon
branching and synaptogenesis, \cmnt{and an assumption that the spatial
  distribution of synaptic density across the cortical sheet is broadly
  homogeneous. As such, the quantity $c_i$ can be thought of as the
  connection density}.

@@

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{The emergence of whisker barrels.
      %
      \textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}.
      %
      \textbf{B} The strengths of interaction $\gamma$ with fields $\rho_1$
      and $\rho_2$ are indicated for each of 41 projections by the lengths of
      green and blue arrows respectively, assuming that similar fields aligned
      to the posterior-anterior and medial-lateral axes in the ventroposterior
      medial nucleus of the thalamus are sampled at the locations of putative
      barreloid centers (reconstructed from \citealp{haidarliu_size_2001},
      their \MPone{Fig.\,5b}).
      %
      \textbf{C} Results for the example simulation, with parameters
      \MPtwo{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$,
        $\gamma\in\pm 2$, $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours
      indicate the thalamic projection for which the connection density is
      maximal, \cmnt{barrel labels are located at the centroid of each
        region and} black lines delineate boundaries (see Movie S1).
      %
      \textbf{D} Red dots show the \emph{Honda}--\metrics{$\delta(t)$} metric
      obtained from \metrics{the} simulation approaching that obtained from
      the \metrics{real} barrels in \textbf{A} (dotted line); black squares
      show the \emph{pattern difference} metric $\eta(t)$, and reveal the
      emergence of a correspondence between the real and simulated barrel
      shapes (units mm$^3$); \metrics{grey hexagons show how selectively each
        cortical site is innervated;
        $\omega(t) = \oiint_{S} \mu(\mb{x},t) \mathrm{d}S$, where
        $\mu(\mb{x}) \equiv \mathrm{max}_i\big(c_i(\mb{x},t)\big)\big/\sum_{j=1}^{N} c_j(\mb{x},t)$.}
      %
      \textbf{E} \metrics{Plotted across the cortical sheet, the selectivity
        develops to reveal an alignment with the emergent barrel boundary
        shapes}. \metrics{Greyscale colour indicates values of $\mu(\mb{x})$.}
      All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

\mpfive{We reasoned that an arbitrary number of distinct field locations may
  be determined by a minimum of two guidance gradients, if the concentration
  of the projection densities is influenced by competition between
  projections, and if a projection that interacts more strongly with a given
  guidance gradient migrates further in the direction of that gradient.
  Accordingly, projections that interact most strongly with a given
  guidance gradient would come to occupy cortical locations at which that
  field has extreme values, leaving adjacent locations available to be
  occupied by projections with the next strongest interactions, and so
  forth. This would in principle allow the \emph{relative} locations of the
  fields to be specified by the relative values of the interaction parameters,
  $\gamma$, and hence for a topological map in the cortex to be specified by a
  spatial ordering of the $\gamma$ values at the level of the thalamus.}

\mpfive{Such dynamics are quite unlike those described by classic
  chemospecificity models} \citep{sperry_chemoaffinity_1963}, \mpfive{which
  essentially assume center-points by specifying conditions in the target
  tissue that instruct pre-identified afferents to stop growing. Consider, for
  example, that when simulated in isolation from one-another, all projections
  in the model described would simply migrate to the extrema of the cortical
  guidance fields.}

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space in which to branch and make connections. The major
  modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{colmptwo}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

\MPone{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).

\MPone{Cells originating in VPM express high levels of EphA receptors and
  project to the lateral part of S1, which expresses low levels of ephrin-A5,
  and cells originating in the VPL express low levels of EphA receptors and
  project to the medial part of S1, which expresses high levels of ephrin-A5}
(see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\MPone{We assume that such patterning arises because the relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

\MPone{For simplicity, the two simulated thalamic interaction gradients, as
  well as the two cortical guidance gradients, were initially chosen to be
  linear and orthogonal. Hence a given pair of $\gamma$ values corresponds to
  the coordinate of a barreloid center in the VPM. Coordinates, in a reference
  plane defined by the anterior-posterior and medial-lateral axes, were
  estimated from Fig 5d of} \cite{haidarliu_size_2001}, \MPone{and scaled
  such that $\gamma\in\pm 2$. Note that this scaling is arbitrary because
  according to the model the coordinates provide relative position information
  only.}

A cortical boundary enclosing barrels for 41 \cmnt{macrovibrissae} was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \mpfour{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$).}

@@

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \metrics{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field
  (for mice $\delta\approx0.054$,} \citealp{senft_mouse_1991}\metrics{, and
  our analysis of data from} \citealp{zheng_signal_2001} \metrics{ indicates that
  the value for rats is similar). For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern can emerge
  within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials \& Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The overall connection selectivity increased as the
  barrel map self-organized in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\MPthreePar{Exploring the parameter space.
        %
        \textbf{A} Colour indicates the quality of the pattern at $t=30000$
        steps, against three measures: Low values of \emph{Honda}--$\delta$
        (top row) suggest a Voronoi-like pattern of fields. The \emph{pattern
          difference}, $\eta$ (middle row) measures the difference in the area
        and arrangement of barrels in real and simulated fields. The
        \emph{selectivity}, $\omega$ (bottom row), measures the specificity
        with which the cortical sheet is innervated. Colour maps are chosen so
        that lighter (orange and yellow) colours indicate higher quality
        patterns against each measure. White squares indicate combinations of
        parameters for which simulations were numerically unstable.
        %
        The parameter space explored is three dimensional with the ratio
        $\alpha/\beta$ varying between plots, and the competition parameter
        $\epsilon$ and the diffusion constant $D$ varying within plots. An
        asterisk (*) marks the parameters used in Fig.\,\ref{fig:main}. Boxes
        i), ii)~and iii)~mark parameter sets for which corresponding patterns
        are shown in \textbf{B} (for $t=30000$ steps).
        %
        \textbf{B} Varying the diffusion constant $D$ generates qualitatively
        different patterns. Higher values cause expansions of the peripheral
        barrels and a corresponding compression of the inner barrels
        (i). Lower values instead cause an expansion of the central barrels
        and compression of the peripheral barrels (ii). Further reducing the
        rate of diffusion (iii), which is equivalent to increasing the size of
        the domain and hence simulating development in an animal with a larger
        cortex, causes a large area to be occupied by projections with
        intermediate interaction parameters; those with strong interaction
        parameters are compressed around the edge of the domain, and
        consequently a barrel pattern fails to form.}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

>>>>

% Black, so that updates following review do not show
\definecolor{colcmnt}        {rgb} {0.0, 0.0, 0.0}
% A command to highlight changes for reviewers
\newcommand{\cmnt}[1]{\textcolor{colcmnt}{#1}}

% Some document-defined commands
\newcommand*\dif{\mathop{}\!\mathrm{d}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}
% Nice norms:
\usepackage{mathtools}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}
% Degrees symbol:
\usepackage{textcomp}

@@

Barrel formation requires afferent input from whisker stimulation and thalamic
calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex
network of axon guidance molecules such as ephrin-A5 and A7 and adhesion
molecules such as cadherin-6 and 8
\citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}.  This network
is orchestrated by interactions between morphogens Fgf8 and Fgf17 and
transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients \cmnt{spanning the cortical sheet} that mark orthogonal axes
and can be manipulated to stretch,
shrink, shift, and even duplicate barrels
\citep{assimacopoulos_fibroblast_2012}.

The barrel boundaries form a Voronoi tessellation \citep{senft_mouse_1991}
(Fig.\,\ref{fig:main}A), suggesting that barreloid topology is preserved in
the projection of thalamocortical axons into the cortex, and that a barrel
forms by lateral axon branching from an initial center-point that ceases upon
contact with axons branching from adjacent centers.  However, the assumption
of pre-arranged center-points is difficult to resolve with the observation
that axons arrive in the cortical plate as an undifferentiated bundle,
\emph{prior} to barreloid formation \citep{agmon_organized_1993}. \cmnt{In
  mice, axons from the trigeminal ganglion arrive in the principal division of
  the trigeminal nucleus (PrV) at E12, then axons from the PrV arrive
  in the ventroposteromedial nucleus of the thalamus (VPM) at E17, then axons
  from the VPM arrive in the cortical plate at E18/P0. Distinct
  whisker-related clusters then become apparent in the PrV at P0-P1, in the
  VPM at P2-P3, and in the cortex at P3-P5}
\citep{erzurumlu_development_2012,sehara_neuronal_2011}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the \cmnt{fraction
  of occupied synapses} $c_i(x,t)$ and \cmnt{the density of} axon branches
$a_i(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for
$N$ thalamocortical projections indexed by $i$. The model was derived from the
assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally
coupled. Extending the original 1D model to simulate arealisation on a 2D
cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total \cmnt{fraction} of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise, \cmnt{$c_i(\mb{x},t)$} increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \cmnt{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). Here $\chi_i=0$ is a
placeholder. The second term on the right represents the coupling between axon
branching and synaptogenesis, \cmnt{and an assumption that the spatial
  distribution of synaptic density across the cortical sheet is broadly
  homogeneous. As such, the quantity $c_i$ can be thought of as the
  connection density}.

@@

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{The emergence of whisker barrels.
      %
      \textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}.
      %
      \textbf{B} The strengths of interaction $\gamma$ with fields $\rho_1$
      and $\rho_2$ are indicated for each of 41 projections by the lengths of
      green and blue arrows respectively, assuming that similar fields aligned
      to the posterior-anterior and medial-lateral axes in the ventroposterior
      medial nucleus of the thalamus are sampled at the locations of putative
      barreloid centers (reconstructed from \citealp{haidarliu_size_2001},
      their \cmnt{Fig.\,5b}).
      %
      \textbf{C} Results for the example simulation, with parameters
      \cmnt{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$,
        $\gamma\in\pm 2$, $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours
      indicate the thalamic projection for which the connection density is
      maximal, \cmnt{barrel labels are located at the centroid of each
        region and} black lines delineate boundaries (see Movie S1).
      %
      \textbf{D} Red dots show the \emph{Honda}--\cmnt{$\delta(t)$} metric
      obtained from \cmnt{the} simulation approaching that obtained from
      the \cmnt{real} barrels in \textbf{A} (dotted line); black squares
      show the \emph{pattern difference} metric $\eta(t)$, and reveal the
      emergence of a correspondence between the real and simulated barrel
      shapes (units mm$^3$); \cmnt{grey hexagons show how selectively each
        cortical site is innervated;
        $\omega(t) = \oiint_{S} \mu(\mb{x},t) \mathrm{d}S$, where
        $\mu(\mb{x}) \equiv \mathrm{max}_i\big(c_i(\mb{x},t)\big)\big/\sum_{j=1}^{N} c_j(\mb{x},t)$.}
      %
      \textbf{E} \cmnt{Plotted across the cortical sheet, the selectivity
        develops to reveal an alignment with the emergent barrel boundary
        shapes}. \cmnt{Greyscale colour indicates values of $\mu(\mb{x})$.}
      All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

\cmnt{We reasoned that an arbitrary number of distinct field locations may
  be determined by a minimum of two guidance gradients, if the concentration
  of the projection densities is influenced by competition between
  projections, and if a projection that interacts more strongly with a given
  guidance gradient migrates further in the direction of that gradient.
  Accordingly, projections that interact most strongly with a given
  guidance gradient would come to occupy cortical locations at which that
  field has extreme values, leaving adjacent locations available to be
  occupied by projections with the next strongest interactions, and so
  forth. This would in principle allow the \emph{relative} locations of the
  fields to be specified by the relative values of the interaction parameters,
  $\gamma$, and hence for a topological map in the cortex to be specified by a
  spatial ordering of the $\gamma$ values at the level of the thalamus.}

\cmnt{Such dynamics are quite unlike those described by classic
  chemospecificity models} \citep{sperry_chemoaffinity_1963}, \cmnt{which
  essentially assume center-points by specifying conditions in the target
  tissue that instruct pre-identified afferents to stop growing. Consider, for
  example, that when simulated in isolation from one-another, all projections
  in the model described would simply migrate to the extrema of the cortical
  guidance fields.}

\cmnt{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space in which to branch and make connections. The major
  modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\cmnt{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{colcmnt}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\cmnt{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

\cmnt{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).

\cmnt{Cells originating in VPM express high levels of EphA receptors and
  project to the lateral part of S1, which expresses low levels of ephrin-A5,
  and cells originating in the VPL express low levels of EphA receptors and
  project to the medial part of S1, which expresses high levels of ephrin-A5}
(see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\cmnt{We assume that such patterning arises because the relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

\cmnt{For simplicity, the two simulated thalamic interaction gradients, as
  well as the two cortical guidance gradients, were initially chosen to be
  linear and orthogonal. Hence a given pair of $\gamma$ values corresponds to
  the coordinate of a barreloid center in the VPM. Coordinates, in a reference
  plane defined by the anterior-posterior and medial-lateral axes, were
  estimated from Fig 5d of} \cite{haidarliu_size_2001}, \cmnt{and scaled
  such that $\gamma\in\pm 2$. Note that this scaling is arbitrary because
  according to the model the coordinates provide relative position information
  only.}

A cortical boundary enclosing barrels for 41 \cmnt{macrovibrissae} was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \cmnt{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$).}

@@

% New section to describe and introduce the metrics
\cmnt{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \cmnt{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field
  (for mice $\delta\approx0.054$,} \citealp{senft_mouse_1991}\cmnt{, and
  our analysis of data from} \citealp{zheng_signal_2001} \cmnt{ indicates that
  the value for rats is similar). For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern can emerge
  within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials \& Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The overall connection selectivity increased as the
  barrel map self-organized in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\cmnt{Exploring the parameter space.
        %
        \textbf{A} Colour indicates the quality of the pattern at $t=30000$
        steps, against three measures: Low values of \emph{Honda}--$\delta$
        (top row) suggest a Voronoi-like pattern of fields. The \emph{pattern
          difference}, $\eta$ (middle row) measures the difference in the area
        and arrangement of barrels in real and simulated fields. The
        \emph{selectivity}, $\omega$ (bottom row), measures the specificity
        with which the cortical sheet is innervated. Colour maps are chosen so
        that lighter (orange and yellow) colours indicate higher quality
        patterns against each measure. White squares indicate combinations of
        parameters for which simulations were numerically unstable.
        %
        The parameter space explored is three dimensional with the ratio
        $\alpha/\beta$ varying between plots, and the competition parameter
        $\epsilon$ and the diffusion constant $D$ varying within plots. An
        asterisk (*) marks the parameters used in Fig.\,\ref{fig:main}. Boxes
        i), ii)~and iii)~mark parameter sets for which corresponding patterns
        are shown in \textbf{B} (for $t=30000$ steps).
        %
        \textbf{B} Varying the diffusion constant $D$ generates qualitatively
        different patterns. Higher values cause expansions of the peripheral
        barrels and a corresponding compression of the inner barrels
        (i). Lower values instead cause an expansion of the central barrels
        and compression of the peripheral barrels (ii). Further reducing the
        rate of diffusion (iii), which is equivalent to increasing the size of
        the domain and hence simulating development in an animal with a larger
        cortex, causes a large area to be occupied by projections with
        intermediate interaction parameters; those with strong interaction
        parameters are compressed around the edge of the domain, and
        consequently a barrel pattern fails to form.}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

\cmnt{Against these three metrics we are also able to characterise the
  robustness of self-organization to the model parameters, and to investigate
  the sensitivity of the model to variation in its inputs.} \cmnt{Fig.\,\ref{fig:paramsweep}A shows values of $\delta$, $\eta$ and
  $\omega$ obtained after 30000 iterations, from 216 independent simulations,
  each representing a unique combination of the model parameters $D$,
  $\epsilon$, and the ratio $\alpha/\beta$. First we observe that
  self-organization is highly robust to the ratio $\alpha/\beta$, across five
  orders of magnitude, with respect to all three metrics. Second, the most
  strongly Dirichletform patterns (low \emph{Honda}-$\delta$) were generated
  by simulations in which the diffusion constant $D$ and the strength of
  competition $\epsilon$ were high. Third, strongest overall connection selectivities,
  $\omega$, were obtained for lower values of $D$. Fourth, variation in the
  pattern difference metric, $\eta$, indicated that the alignment between real
  and simulated patterns was greatest for intermediate rates of diffusion,
  $D\approx0.5$. Together these results indicate that when competition is
  strong, the rate of diffusion determines a trade-off such that fields emerge
  to be barrel-shaped when diffusion is fast and they emerge to be more
  selectively innervated when diffusion is slow.}

\cmnt{The parameters of the example simulation are indicated in
  Fig.\,\ref{fig:paramsweep}A using an asterisk. In
  Fig.\,\ref{fig:paramsweep}B, we also present examples of alternative
  patterns that emerge for different choices of $D$. Decreasing the rate of
  diffusion may be considered equivalent to increasing the overall size of the
  domain, $S$. Hence, insights into barrel development in species with a
  larger representation of the vibrissae, which do not have barrel fields, may
  be gained by studying pattern formation
  when $D$ is small. In this context, it is interesting to note that for small
  $D$, the organization is predicted to be topological but highly irregular,
  with a general expansion in the territory occupied by the central versus
  peripheral domains that would presumably manifest as an absence of
  identifiable barrel fields (Fig.\,\ref{fig:paramsweep}B\,iii).}
